Historical collections as baseline for assessing the extent of the pollinator

### 2018-04-05 ms.Rmd
Small edits, too tired to finish it today :D

[Historical collections or museum collections?]()

@@

There is increasing concern about the decline of pollinators worldwide. However, despite reports that pollinator declines are widespread, data are scarce and often geographically and taxonomically biased. These biases limit robust inference about any potential “pollinator crisis”. Analysis of historical time series data is the most direct approach for assessing species declines, but long term monitoring programs targeting pollinators are limited. However, non-structured and often opportunistic historical data collections are more frequently available. In many instances, these data provide the only source of historical information serving as a baseline for identifying pollinator declines. Specimens historically collected and preserved in museums can provide information on where and when species were collected, in addition to other ecological information (e.g., historical species interactions and morphological traits). Here, we provide a synthesis of how researchers have used museum specimens to compare historical occurrences with current patterns to identify long-term changes in biodiversity, species abundances, body size and pollination services. Despite recent advances, we show that information on the status and trends of most pollinators is absent. Furthermore, we identify lack of collection record data digitization as a key bottleneck, impeding progress in this field. We highlight opportunities and limitations to progress assessment of pollinator declines globally. Finally, we demonstrate different approaches to analysing museum collection data using two contrasting case studies from distinct geographical regions (New Zealand and Spain) for which long-term pollinator declines have never been assessed. We show that for both regions, winner and loser species emerge despite non-significant changes in species richness over time. There is immense potential for museum specimens to play a central role in assessing the extent of the global pollination crisis, but to progress this field data must be made feely accessible to researchers.

@@

One solution to overcoming these barriers has been to use space-for-time substitutions, where researchers compare pollinator populations across environmental gradients. Despite critiques on the robustness of this approach [@Franca2016; @DePalma2018], these studies currently provide the most extensive source of data on pollinator populations. For example, researchers have recently estimated bee richness declines for every country in Europe using predictions from models of pollination associations with different land-use types [@DePalma2017]. A second important method is the use of data collected from pollinator monitoring programs, which are often driven by citizen scientists. This approach was inspired by successful butterfly monitoring programs [@vanSwaay2008] and is currently being extended to other pollinator taxa. However, these programs require significant time to generate long-term datasets and cannot be used to assess historic pollinator populations. Finally, the most practical approach for assessing long-term historical pollinator population trends is to use the historical information on species occurrences, which is archived in museum collections [e.g., @Bartomeus2013a].

In this review, we first assess current evidence for pollinator richness declines and present a road map outlining a strategy for using museum collection data to fill current knowledge gaps. We highlight the major technical difficulties involved in using museum collection data and demonstrate several approaches for analysing different types of museum collection data to assess long-term trends in pollinator populations. Finally, we highlight the need to move beyond simple biological diversity descriptors and unleash the power of historical data to assess changes in species interactions, ecosystem functioning and evolutionary changes through time.

@@

#Using historical museum specimen records to fill knowledge gaps

As shown in Figure 1, specific estimates are lacking for countries worldwide and use of historical museum collection data may be the most effective tool for filling those gaps.

@@

Once museum collection datasets are made available, researchers must identify any potential biases in the data. We explore this process with two contrasting dataset examples (Spain and New Zealand). In the Spanish dataset, most of the data comes from a few specific locations and was collected by the same team of people. In this case, we contacted the original collectors to define their sampling protocols. We then resurveyed the same sites (20 years after the original surveys) using the same protocols. In contrast, the New Zealand dataset includes a wide suite of collectors and collection locations but shows no obvious biases in geographical and taxonomic coverage through time. For these case studies, we provide annotated R scripts for analysis of different dataset types (e.g., for Spain and New Zealand, as described above). These different analytical approaches allow us to reveal long-term trends in pollinator populations for regions with contrasting sampling histories. We hope this resource will encourage researchers to analyse data for regions where there is current information on pollinator declines is lacking.

>>>>

@@

There is increasing concern about the decline of pollinators worldwide. However, despite reports that pollinator declines are widespread, data are scarce and often geographically and taxonomically biased. These biases limit robust inference about any potential “pollinator crisis”. Analysis of historical time series data is the most direct approach for assessing species declines, but long term monitoring programs targeting pollinators are limited. However, non-structured and often opportunistic historical data collections are more frequently available. In many instances, these data provide the only source of historical information serving as a baseline for identifying pollinator declines. Specimens historically collected and preserved in museums can provide information on where and when species were collected, in addition to other ecological information (e.g., historical species interactions and morphological traits). Here, we provide a synthesis of how researchers have used museum specimens to compare historical occurrences with current patterns to identify long-term changes in biodiversity, species abundances, body size and pollination services. Despite recent advances, we show that information on the status and trends of most pollinators is absent. Furthermore, we identify lack of collection record data digitization as a key bottleneck, impeding progress in this field. We highlight opportunities and limitations to progress assessment of pollinator declines globally. Finally, we demonstrate different approaches to analysing museum collection data using two contrasting case studies from distinct geographical regions (New Zealand and Spain) for which long-term pollinator declines have never been assessed. There is immense potential for museum specimens to play a central role in assessing the extent of the global pollination crisis, but to progress this field data must be made feely accessible to researchers.

@@

One solution to overcoming these barriers has been to use space-for-time substitutions, where researchers compare pollinator populations across environmental gradients. Despite critiques on the robustness of this approach [@Franca2016; @DePalma2018], these studies currently provide the most extensive source of data on pollinator populations. For example, researchers have recently estimated bee richness declines for every country in Europe using predictions from models of pollination associations with different land-use types [@DePalma2017]. A second important method is the use of data collected from pollinator monitoring programs, which are often driven by citizen scientists. This approach was inspired by successful butterfly monitoring programs [@vanSwaay2008] and is currently being extended to other pollinator taxa. However, these programs require significant time to generate long-term datasets and cannot be used to assess historic pollinator populations. Finally, the most practical approach for assessing long-term historical pollinator population trends is to use the historical information on species occurrences, which is often archived in museum collections [e.g., @Bartomeus2013a].

In this review, we first assess current evidence for pollinator richness declines and present a road map outlining a strategy for using historical collection data to fill current knowledge gaps. We highlight the major technical difficulties involved in using historical collection data and demonstrate several approaches for analysing different types of historical collection data to assess long-term trends in pollinator populations. Finally, we highlight the need to move beyond simple biological diversity descriptors and unleash the power of historical data to assess changes in species interactions, ecosystem functioning and evolutionary changes through time.

@@

#Using historical collection specimen records to fill knowledge gaps

As shown in Figure 1, specific estimates are lacking for countries worldwide and use of historical  collection data deposited in museums may be the most effective tool for filling those gaps.

@@

Once museum collection datasets are made available, researchers must identify any potential biases in the data. [Talk about the caveats of teh above crude analysis!]() We explore this process with two contrasting dataset examples (Spain and New Zealand). In the Spanish dataset, most of the data comes from a few specific locations and was collected by the same team of people. In this case, we contacted the original collectors to define their sampling protocols. We then resurveyed the same sites (20 years after the original surveys) using the same protocols. In contrast, the New Zealand dataset includes a wide suite of collectors and collection locations but shows no obvious biases in geographical and taxonomic coverage through time. For these case studies, we provide annotated R scripts for analysis of different dataset types (e.g., for Spain and New Zealand, as described above). These different analytical approaches allow us to reveal long-term trends in pollinator populations for regions with contrasting sampling histories. We hope this resource will encourage researchers to analyse data for regions where there is current information on pollinator declines is lacking.
